Live and Let Die? (Delayed Issues not Harder to Resolve)

### 2015-08-14 data.tex
Updates from ShareLaTeX

The phases logged by our data are shown in \fig{waterfall}.
 Although the representation suggests a waterfall model, the SEI experience is that all real implementations of any size follow a spiral approach with many team performing the work in iterative and/or incremental development cycles. %Within a project increment, multiple features or components may be developed and incremented. TSP is compatible %with agile and encourages iterative and incremental development. Nonetheless, the specific strategy and cycle %duration is a project decision. TSP does, however, strongly encourage 1) constructing units of sufficient size that %%measurement is practicable, and 2) separating the construction from appraisal and test activities. This %effectively highlights the separation of construction from rework activities and aids the apportionment of defects %to those found in appraisal activities (reviews and inspections) and those found through failure (test). The %distinct construction activities (requirements, high and detailed design, and code) were chosen to help teams %analyze the effectiveness and efficiency of their practices through analysis of the defect phase origin, type, fix %effort, and phase of discovery.

>>>>

The phases logged by our data are shown in \fig{waterfall}.
 Although the representation suggests a waterfall model, the SEI experience is that all real implementations of any size follow a spiral approach with many team performing the work in iterative and/or incremental development cycles.

One special feature of  \fig{waterfall} is
the {\em before} where the   TSP Launch Preparation scripts and checklists (see TSP: Leading a development Team~\cite{Humphrey:2005}) require the TSP team to assure that management and business have clearly identified cost, schedule, and scope goals appropriate to the upcoming development activities, often including a conceptual model. For example an architecture team must have sufficient requirements to reason about, prototype, and specify an architecture~\cite{Bachmann13} while a coding only team within a larger project would have more precisely defined requirements and high level design.

%Within a project increment, multiple features or components may be developed and incremented. TSP is compatible %with agile and encourages iterative and incremental development. Nonetheless, the specific strategy and cycle %duration is a project decision. TSP does, however, strongly encourage 1) constructing units of sufficient size that %%measurement is practicable, and 2) separating the construction from appraisal and test activities. This %effectively highlights the separation of construction from rework activities and aids the apportionment of defects %to those found in appraisal activities (reviews and inspections) and those found through failure (test). The %distinct construction activities (requirements, high and detailed design, and code) were chosen to help teams %analyze the effectiveness and efficiency of their practices through analysis of the defect phase origin, type, fix %effort, and phase of discovery.

### 2015-08-14 phaseDelay.tex
Updates from ShareLaTeX

In summary, there exists (1)~historical reasons for believing in the delayed issued
effect as well as (2)~reasons to believe that that effect may not longer hold.
Hence,  this article explores the currency of the delayed issue effect.
After some initial definitions, we discuss the
value of checking old ideas. Next, survey of industrial practitioners and researchers to document a widespread belief that delayed issues have a negative impact on projects.  Next, we  analyze 171 software  projects developed in the period 2006 to 2014 to find  {\em no evidence} of the delayed issue effect
(note that, to ensure reproducibility,
all the data  used in this study is available in the PROMISE
repository at openscience.us/repo). Finally, we discuss the validity and implications of our results.

\section{Definitions \& Hypotheses}
This paper uses the following definitions:
\bi
\item
{\em Delayed issue effect}:   it is {\em exponentially}  more {\em difficult} to resolve an issue in a software project, the {\em longer} it remains.
\item
 {\em Longer} time is defined as per  Boehm'81~\cite{Boehm81}; i.e. the gap between the   phases where   issues are introduced and resolved.
\item
A measure collected in phase ${1,.,i,..j}$ is 
{\em exponential} when  that
   measure at phase $j$   
   is larger than the sum of that measure in 
earlier phases $1 \le i < j$. 
\item
Issues are more {\em difficult}  
when their resolution takes more time or costs more  (e.g. needs expensive
debugging tools or the skills of expensive developers).
\ei
Note that we use  the  term ``delayed issue effect'' as generalization of the
more specific rule  ``requirements errors are hardest to fix''.
This generalization is valid since the rationale for the rule about requirements
is usually done as per McConnell~\cite{mcconnell01}; i.e. small upstream mistakes very
early in the system can cascade into huge problems later in the lifecycle.
That said, we prefer our more general term ``delayed issue effect'' since not
only might requirements errors cascade, so too might analysis errors, design errors, etc.

This paper defects the  following claims. Note that we call them
``claims'' not ``hypotheses'' since the later require defense via some statistical
significance test. On the other hand, our claims will be supported via a variety
of arguments, presented later in this paper.

@@

\section{``DIE'' is a  widely-held belief}
To assess the prevalence of the delayed issue effect,   conducted a survey of software engineers. If our surveyed practitioners make management decisions based on their
understanding of SE theory, then the phase delay effect may well influence their judgment.

@@

Using Figure~\ref{fig:fix-phase-dist}, we can comment on two specific and more general
delayed issued effects. 
\bi
\item {\bf Specific1}: Requirements errors are the most expensive to fix;
\item {\bf Specific2}: Requirements errors are the most expensive to fix, 
 when found during production but the cheapest to fix early in development;
 \item {\bf General:}  It is exponentially more  difficult  to resolve an issue in a software project, the   longer it remains.
\ei

\subsubsection{Specific1}

Our data does not support   
{\em  requirements errors are the most expensive to fix}.
Note that   median time spent fixing errors from {\em requirements} across the life cycle was 92, 93,  49, 33, 39, 49, 67, 122 minutes\footnote{If this time seems long, then recall
that it includes the time required  to (a)~collect data and realize there is an error;
(b)~prepare a fix;  and (c)~apply some validation
procedure to check the fix (e.g. discuss it with a colleague or execute some tests).}.
This is:
\bi
\item
Clearly longer than median times seen {\em Before } and in {\em Planning}
(30,28,45,77,65,56,24,35,33);
\item
Somewhat shorter, but still not exponentially smaller,  than the median times see after {\em requirements} in {\em design}
(103,101, 79,66, 119,88, 80,45) and in {\em code} (65,115, 113,125, 95,133). 
\ei
For completeness, we ran an  
A12 test comparing the raw data (and not just the medians of \fig{raw}) and found
 a non-small effect only for the difference {\em requirements}
to {\em code}. But even though the effect was non-small, 
 the overall size of the difference
{\em requirements} to {\em code} was less than 30\% (i.e. much less than we would
expect from the delayed issue effect). 
\subsubsection{Specific2}
Next, we comment on 
requirements errors are the most expensive to fix 
{\em when found during production but the cheapest to fix early in development}. 
Our data from these 171 project does not extend into post-delivery deployment. That said,
there is enough evidence to raise doubts that our projects will exhibit exponentially more
difficult issues after deployment. 
Specifically: prior reports of an  exponential increase in difficulty in resolving issues {\em during
deployment}.
For example:
\bi
\item  From requirements to acceptance testing
\fig{b81} reports a 40-fold increase;
\item
 \fig{cost-to-fix} reports a 100-fold increase
(for the larger projects).
\item
The projects surveyed in \fig{steck} report changes of a  \newline
\{37,40,40,50,51,70,75,250\}-fold increase in pre-deployment period when the software was being developed.
\ei
Any manager watching  these intra-development trends would be well-justified
to anticipate that  the difficulty in resolving issues would will get much worse. Such managers
would therefore expect an exponentially harder time resolving issues during deployment.

On the other hand, a remarkable feature of Figure~\ref{fig:fix-phase-dist} is the {\em absence}
of an exponential delay effect during development.
In our data, the greatest difficulty in resolving requirements issues was 1.31 seen in systems testing--
which is far smaller than the 37 to 250-fold increases reported in \fig{b81}, \fig{cost-to-fix},
and \fig{steck}.  That is, from our  data, we would not anticipate exponentially greater difficulty in resolving requirements issues after
deployment. 
 \subsubsection{General}
 The above specific statements related just to requirements. Turning now
 to issues creased in all phases in Figure~\ref{fig:fix-phase-dist},  we see that
 this data does not follow the exponential increase predicted
by the delayed issue effect:
\bi
\item There is much ``up and down'' in the chart. 
\item If we separate out just the {\em Inspect} or {\em Review} data, we can see that there is some increase in {\em Design} and {\em Code}. However, that effect could be a linear or be just noise. In either case,
it does not look like exponential.
\item The time in final testing is particularly low suggesting that few defects survived into testing
(and hence, there were few instances where code required massive rework).
\ei
In summary:
\bi
\item
The delayed issue effect has been reported in around a dozen papers, at least half of which date back to the 1970s. 
\item
But there is no trace of that  effect   in hundreds of software projects built  since 2006.
\ei
%For illustrative purposes, the relative cost to fix errors per lifecycle phase in the TSP project data is overlayed with Boehm's cost curve~\cite{Boehm81} as well as the CCPDS-R case study~\cite{Royce98} in \fig{cost-to-fix-tsp}. Cost-to-fix for TSP is total time spent in each review, inspection, or test-related phase / \# of defects found in phase (reported in \fig{raw}). It is clear to see that the trend is relatively flat, closer to the pattern exhibited by the Royce case study (which also focused on early defect removal) than the exponential growth curve.

@@

>>>>

The above argument is an anecdotal evidence that  the delayed issue effect might no longer exist. But anecdotes are not really rigorous
evidence. Accordingly,  this article explores the currency of the delayed issue effect.
After some initial definitions, we discuss the
value of checking old ideas. Next, a survey of industrial practitioners and researchers to document a widespread belief that delayed issues have a negative impact on projects.  After that, we  analyze 171 software  projects developed in the period 2006 to 2014 to find  {\em no evidence} of the delayed issue effect:
\bi
\item
To ensure reproducibility,
all the data  used in this study is available in the PROMISE
repository at openscience.us/repo. 
\item
To the best of our knowledge,
this the largest study devoted the delayed issue effect yet conducted.
\ei
Finally, we discuss the validity and implications of our results.

\section{Definitions \& Hypotheses}
This paper uses the following definitions:
\bi
\item
The {\em delayed issue effect}:   it is {\em exponentially}  more {\em difficult} to resolve  issues in a software project, the {\em longer} they remain.
\item
 {\em Longer} time is defined as per  Boehm'81~\cite{Boehm81}; i.e. the gap between the   phases where   issues are introduced and resolved.
\item
A measure collected in phase ${1,.,i,..j}$ is 
{\em exponential} when  that
   measure at phase $j$   
   is larger than the sum of that measure in 
earlier phases $1 \le i < j$. 
\item
Issues are more {\em difficult}  
when their resolution takes more time or costs more  (e.g. needs expensive
debugging tools or the skills of expensive developers).
\ei
Note that we use  the  term ``delayed issue effect'' as generalization of the
more specific rule  ``requirements errors are hardest to fix''.
This generalization is valid since the rationale for the rule about requirements
is usually done as per McConnell~\cite{mcconnell01}; i.e. small upstream mistakes very
early in the system can cascade into huge problems later in the lifecycle.
That said, we prefer our more general term ``delayed issue effect'' since not
only might requirements errors cascade, so too might analysis errors, design errors, etc.

This paper defends the  following claims. Note that we call them
``claims'' not ``hypotheses'' since the later require defense via some statistical
significance test. On the other hand, our claims will be supported via a variety
of arguments, presented later in this paper.

@@

\section{``DIE'' is a  widely-held belief}
To assess the prevalence of DIE,   we conducted a survey of software engineers. If our surveyed practitioners make management decisions based on their
understanding of SE theory, then the DIE  may well inappropriately influence their decisions.

@@

\subsubsection{Comment 1}
Figure~\ref{fig:fix-phase-dist}
 data does not support   
{\em  requirements errors are the most expensive to fix}.
Note that   median time spent fixing errors from {\em requirements} across the life cycle was 92, 93,  49, 33, 39, 49, 67, 122 minutes\footnote{If this time seems long, then recall
that it includes the time required  to (a)~collect data and realize there is an error;
(b)~prepare a fix;  and (c)~apply some validation
procedure to check the fix (e.g. discuss it with a colleague or execute some tests).}.
This is:
\bi
\item
Clearly longer than median times seen {\em Before } and in {\em Planning}
(30,28,45,77,65,56,24,35,33);
\item
Somewhat shorter, but still not exponentially smaller,  than the median times see after {\em requirements} in {\em design}
(103,101, 79,66, 119,88, 80,45) and in {\em code} (65,115, 113,125, 95,133). 
\ei
For completeness, we ran an  
A12 test comparing the raw data (and not just the medians of \fig{raw}) and found
 a non-small effect only for the difference {\em requirements}
to {\em code}. But even though the effect was non-small, 
 the overall size of the difference
{\em requirements} to {\em code} was less than 30\% (i.e. much less than we would
expect from the delayed issue effect).

That said, Figure~\ref{fig:fix-phase-dist} does support the 
claim that early lifecycle errors are most expensive to fix.
In all our results, the biggest scale ups are found with resolving
issues introduced in the {\em before} phase. For example,
resolving {\em before} issues during unit testing was 2.56 times
slower than resolving issues earlier in the lifecycle.

Nevertheless,  it must be added that the scale up seen in 
Figure~\ref{fig:fix-phase-dist} 
is much less
than suggested by the DIE literature. Returning to
the {\em before/unit test} result: 2.56 is far smaller
than the scale ups seen in Figure~1. In fact, nowhere in 
our results do we see any exponential increase.

\subsubsection{Comment 2}
Next, we comment on 
requirements errors are the most expensive to fix 
{\em when found during production but the cheapest to fix early in development}. 
Our data from these 171 project does not extend into post-delivery deployment. That said,
there is enough evidence to raise doubts that our projects will exhibit exponentially more
difficult issues after deployment.

That evidence is as follows.
Prior reports of an  exponential increase in difficulty in resolving issues {\em during
development}.
For example:
\bi
\item  From requirements to acceptance testing
\fig{b81} reports a 40-fold increase;
\item
 \fig{cost-to-fix} reports a 100-fold increase
(for the larger projects).
\item
The projects surveyed in \fig{steck} report changes of a  \newline
\{37,40,40,50,51,70,75,250\}-fold increase in pre-deployment period when the software was being developed.
\ei
Any manager watching  these intra-development trends would be well-justified
to anticipate that  the difficulty in resolving issues would will get much worse. Such managers
would therefore expect an exponentially harder time resolving issues during deployment.

On the other hand, a remarkable feature of Figure~\ref{fig:fix-phase-dist} is the {\em absence}
of an exponential delay effect during development.
In our data, the greatest difficulty in resolving requirements issues was 1.31 seen in systems testing--
which is far smaller than the 37 to 250-fold increases reported in \fig{b81}, \fig{cost-to-fix},
and \fig{steck}.  That is, from our  data, we would not anticipate exponentially greater difficulty in resolving requirements issues after
deployment. 
 \subsubsection{Comment 3}
 The above specific statements related just to requirements. Turning now
 to issues creased in all phases in Figure~\ref{fig:fix-phase-dist},  we see that
 this data does not follow the exponential increase predicted
by the delayed issue effect:
\bi
\item There is much ``up and down'' in the chart. 
\item If we separate out just the {\em Inspect} or {\em Review} data, we can see that there is some increase in {\em Design} and {\em Code}. However, that effect could be a linear or be just noise. In either case,
it does not look like exponential.
\item The time in final testing is particularly low suggesting that few defects survived into testing
(and hence, there were few instances where code required massive rework).
\ei
In summary:
\bi
\item
The delayed issue effect has been reported in around a dozen papers, at least half of which date back to the 1970s. 
\item
But there is no trace of that  effect   in hundreds of software projects built  since 2006.
\ei
%For illustrative purposes, the relative cost to fix errors per lifecycle phase in the TSP project data is overlayed with Boehm's cost curve~\cite{Boehm81} as well as the CCPDS-R case study~\cite{Royce98} in \fig{cost-to-fix-tsp}. Cost-to-fix for TSP is total time spent in each review, inspection, or test-related phase / \# of defects found in phase (reported in \fig{raw}). It is clear to see that the trend is relatively flat, closer to the pattern exhibited by the Royce case study (which also focused on early defect removal) than the exponential growth curve.

@@

Another issue that threatens internal validity is that our
results are dependent on developers correctly identifying which
phase initially created an issue. Certainly, if that was done
incorrectly in this study, then all our results must be questioned.
However,  this issue threatens {\em every} study on the delayed
issue effect-- so if our results are to be doubted on this score,
then all prior work that reported the delayed issue effect should
also be doubted. Moreover, the TSP method used in this study
encourages greater care with error reporting:

\bi
\item Developers are trained   to make that judgement;
\item All data entry is double-checked by the team couch 
\item TSP demands that developers analyze their data to make
process improvements. 
\ei
That is, TSP developers are always testing
if their project insights are accurate. In such an environment,
it is more likely that they will better identify the injection phase.
